---
title: "Visualizing IMDB ratings"

output:
  html_document:
    toc: true
    toc_float: true
    includes:
      in_header: 
        header.html
navbar:
  title: "Home"
  left:
    - text: "About Me"
      href: about_me.html
    - text: "Portfolio"
      menu:
        - text: "Classifying wine quality"
          href: FinalProject.html
        - text: "Visualizing IMDB ratings"
          href: imdb.html
    - text: "Contact Me"
      menu:
        - text: "jsteven.raquel8@gmail.com"
        - text: Twitter
          href: https://twitter.com/jsteven_r
        - text: GitHub
          href: https://github.com/jstevenr
        - text: LinkedIn
          href: https://www.linkedin.com/in/jstevenr/
        - text: Medium
          href: https://medium.com/@jstevenr          
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

----

This was a very simple visualization I did to test out some of the abilities of `ggplot2`, an `R` package with some more robust options that go beyond the base level visualization abilities of `R`.  

We seek to visualize movies over the past few decades to see how quality (or perceived quality at least) has changed over time. There is no actual analysis done on this data, save for the kind of intuitive deduction that can be made from observing the visualizations. Having said that, I would be interested in delving more deeply into these data with the skills in data mining that I've gained since having done this.

Data was found on [Kaggle.com](https://www.kaggle.com/deepmatrix/imdb-5000-movie-dataset) and drawn from the [Internet Movie Database](imdb.com). The code to create the plots can be found in the accompanying GitHub [repo](https://github.com/jstevenr/imdb-ratings) for this project, as `imdb.Rmd`.

----

```{r, include = FALSE}
library(ggplot2)
library(gridExtra)
library(readr)
```

# Data pre-processing

Here we `subset()` the data in order to work with a data that makes sense for our purposes. I decided that it would be good to only look at films that were voted on by at least 10,000 people, to ensure that these ratings would be more accurate. For example, a universally praised (or panned) film that only has 1000 votes, probably because it was only seen by a very small minority, probably shouldn't factor into our appraisal of the the quality of films of the decade in general.

We're also only going to be looking at the 1980s to the present, so we drop the other rows, as well as any columns that don't include what we are looking for.


```{r, data cleaning}
all_movies <- read.csv("~/Documents/git-workspace/imdb-ratings/movie_metadata.csv",
                   stringsAsFactors = FALSE, 
                   na.strings = c(NA, "NA", "", 0, "0"), 
                   skipNul = TRUE,
                   fill = FALSE)

# Only retain films with sufficient number of votes
all_movies <- subset(all_movies, num_voted_users >= 20000, 
                 select = c(imdb_score, title_year, movie_title, director_name, gross),                  drop = TRUE)

# To remove any rows with missing values
all_movies <- all_movies[complete.cases(all_movies),]

# Remove any duplicate films
all_movies <- all_movies[!duplicated(all_movies$movie_title),]

stats <- matrix(data = NA, nrow = 6, ncol = 7, 
                dimnames = list(c("1927-2016", "1980-2016","2010-2016", "2000-2010", "1990-2000","1980-1990"),
                                c("Min", "1Q", "Med", "Mean", "3Q", "Max", "Size")))

# Adding summary statistics and size
stats[1,1:6] <- round(summary(all_movies$imdb_score), 2)
stats[1,7] <- nrow(all_movies)

all_avg <- stats[1,4]
```

The films range from scoring between 1.6 and 9.3 on a scale from 0 to 10.0m and the average score is about 6.62.


```{r, subsetting data}
# Keeping only relevant columns as well as only looking at movies from 1980s and beyond
movies <- subset(all_movies, title_year >= 1980)

stats[2,1:6] <- round(summary(movies$imdb_score), 2)
stats[2,7] <- nrow(movies)
modern_avg <- stats[2,4]
stats
```

In the process of dropping films before 1980, the average scores were scarcely affected. This is most likely because of the fact that neither IMDB nor the internet existed since the earliest days of film. Knowing this, rather than look at the average score across the entirety of film we're only going to look at the average between 1980-2016, and compare these films with that metric. 

----

# The 10s

The data for the decade that began in 2010 only goes up until 2016, so naturally it is a smaller subset of data than the rest. 

Let's take a look at the best and worst films of the decade.

```{r, organizing the 10s}
tens <- subset(movies, title_year >= 2010)
tens <- tens[order(tens$imdb_score, decreasing = TRUE),]

stats[3,1:6]<- round(summary(tens$imdb_score), 2)
stats[3,7] <- nrow(tens)
stats
```

Overall, the films from 2010-2016 are roughly similar in quality to those across all years and after 1980. The best film isn't here though, although the worst film might be, judging by the mins and maxes. 

```{r}
head(tens)
```

Note the fact that the three best films of this 6-year period (as voted by IMDB users) were all by Christopher Nolan. Also tied for third is my favorite director, Quentin Tarantino with *Django Unchained*. *Samsara*, a relatively unknown film, grossed a modest $2.6 million, a mere fraction of the other films on this list.

```{r, tens tail}
tail(tens)
```

On the other side of things, the worst film of this period was apparently Justin Bieber's concert film, *Never Say Never*. Two parody films topped the list. This film, as well as *Jack and Jill*, starring Adam Sandler, grossed well over $70 million, most likely owing to their leads, but nevertheless were universally panned by audiences. 

We can also take a look at the averages across the individual years as well as each decade.


```{r, 2010-present day, echo = FALSE}
tens_plot <- ggplot(tens, aes(x = factor(title_year), y = imdb_score)) +
  geom_boxplot(fill = "#4271AE", colour = "#1F3552") +
  theme(legend.position = "none") +
  theme(axis.text.x = element_text(hjust = 1)) + 
  scale_y_continuous(breaks = seq(0,10,1), name = "IMDB Rating") +
  scale_x_discrete(breaks = seq(2010, 2016, 1), name = "Year") + 
  ggtitle("2010s in film as rated by IMDB users") +
  theme_bw() + stat_boxplot(geom = 'errorbar') + geom_jitter() +
  geom_hline(aes(yintercept = modern_avg, colour = "Average IMDB Rating from 1980-2016"), show.legend = TRUE) +
  theme(legend.position = "bottom")
tens_plot
```

Modern films seem to be pretty mediocre across the board (between 6-7), but if nothing else, each year seems to be consistent with the last. The sole high outlier of this decade was the 2010 film *Inception*, directed by Christopher Nolan. 


----

# The 00s

```{r, organizing the 00s}
zeroes <- subset(movies, title_year >= 2000 & title_year < 2010)
zeroes <- zeroes[order(zeroes$imdb_score, decreasing = TRUE),]

stats[4,1:6]<- round(summary(zeroes$imdb_score), 2)
stats[4,7] <- nrow(zeroes)
stats
```

```{r}
head(zeroes)
```

All three of Peter Jackson's *The Lord of the Rings* trilogy cracked the top 6, along with another Nolan Batman film. Two foreign films made the top of this list, the Brazilian film *City of God* (aka *Cidade de Deus*) and a personal favorite, along with the animated Miyazaki classic, *Spirited Away*.

```{r, zeroes tail}
tail(zeroes)
```

Notably, two of the worst films of this decade were by Jason Friedberg, namely the parody films *Epic Movie* and *Disaster Movie*. He is also responsible for *Vampires Suck*, one of the worst rated films of the 2010s. 


```{r, 2000-2009, echo = FALSE}
zeroes_plot <- ggplot(zeroes, aes(x = factor(title_year), y = imdb_score)) +
  geom_boxplot(fill = "#4271AE", colour = "#1F3552") +
  theme(axis.text.x = element_text(hjust = 1)) +
  scale_y_continuous(breaks = seq(0,10,1), name = "IMDB Rating") +
  scale_x_discrete(breaks = seq(2000, 2009, 1), name = "Year") + 
  ggtitle("2000s in film as rated by IMDB users") +
  theme_bw() + stat_boxplot(geom = 'errorbar') + geom_jitter() +
  geom_hline(aes(yintercept = modern_avg, colour = "Average IMDB Rating from 1980-2016"), show.legend = TRUE) +
  theme(legend.position = "bottom")
zeroes_plot
```

----

# The 90s

```{r, organizing the nineties}
nineties <- subset(movies, title_year >= 1990 & title_year < 2000)
nineties <- nineties[order(nineties$imdb_score, decreasing = TRUE),]

stats[5,1:6]<- round(summary(nineties$imdb_score), 2)
stats[5,7] <- nrow(nineties)
stats
```

```{r}
head(nineties)
```

Three of my favorite films came out in the 1990s and topped this list: *Pulp Fiction*, *Fight Club*, and *The Matrix*. The decade's top film, *The Shawshank Redemption*, actually was a box office disappointment because of its direct competition with Tarantino's film.

```{r}
tail(nineties)
```

Two films based on fighting video games (my personal favorite genre) landed in the bottom spots: *Street Fighter*, notably starring Jean Claude van Damme and Kylie Minogue, and *Mortal Kombat: Annihilation*. Also included was the only Batman film starring George Clooney, replacing Val Kilmer. It would be the last Batman film until Nolan's reboot in 2005 (the third of his Batman trilogy of course topping the charts in the 2010s). Arnold Schwarzeneggar was Mr. Freeze in this film, possibly the 'punniest' villain of all time. 

```{r, 1990-1999, echo = FALSE}
nineties_plot <- ggplot(nineties, aes(x = factor(title_year), y = imdb_score)) +
  geom_boxplot(fill = "#4271AE", colour = "#1F3552") +
  theme(legend.position = "none") +
  theme(axis.text.x = element_text(hjust = 1)) + 
  scale_y_continuous(breaks = seq(0,10,1), name = "IMDB Rating") +
  scale_x_discrete(breaks = seq(1990, 1999, 1), name = "Year") + 
  ggtitle("1990s in film as rated by IMDB users") +
  theme_bw() + stat_boxplot(geom = 'errorbar') + geom_jitter() +
  geom_hline(aes(yintercept = modern_avg, colour = "Average IMDB Rating from 1980-2016"), show.legend = TRUE) +
  theme(legend.position = "bottom")
nineties_plot
```



----

# The 80s

```{r}
eighties <- subset(movies, title_year >= 1980 & title_year < 1990)
eighties <- eighties[order(eighties$imdb_score, decreasing = TRUE),]

stats[6,1:6] <- round(summary(eighties$imdb_score), 2)
stats[6,7] <- nrow(eighties)
stats
```

```{r}
head(eighties)
```

Robert Zemeckis, who made the last list in the 90s for *Forrest Gump*, also topped this decade with one of my favorites, *Back to the Future*. The two latter films in the original Star Wars trilogy, Episodes V and VI, also topped the list.

```{r, eighties tail}
tail(eighties)
```

Speaking of sequels however, the worst films of the 80s were several ill-advised sequels to well-known classics, many of which were done without the involvement of original directors or cast members. Of the films here, only *Howard the Duck* is not a sequel. Notably also is the inclusion of the 5th and 8th *Friday the 13th* films. *Halloween III* is the also only film in the series that doesn't not feature the antagonist Michael Myers. 

*Jaws: The Revenge* was the fourth and final sequel to the series, with only the original film involving Steven Spielberg, who of course directed the 2nd best rated film of this decade, *Raiders of the Lost Ark* as well as *Schindler's List* in the 90s.


```{r, 1980-1989, echo = FALSE}
eighties_plot <- ggplot(eighties, aes(x = factor(title_year), y = imdb_score)) +
  geom_boxplot(fill = "#4271AE", colour = "#1F3552") +
  theme(legend.position = "none") +
  theme(axis.text.x = element_text(hjust = 1)) + 
  scale_y_continuous(breaks = seq(0,10,1), name = "IMDB Rating") +
  scale_x_discrete(breaks = seq(1980, 1989, 1), name = "Year") + 
  ggtitle("1980s in film as rated by IMDB users") +
  theme_bw() + stat_boxplot(geom = 'errorbar') + geom_jitter() + 
  geom_hline(aes(yintercept = modern_avg, colour = "Average IMDB Rating from 1980-2016"), show.legend = TRUE) +
  theme(legend.position = "bottom")
eighties_plot
```

----

# Conclusions

```{r, echo = FALSE}
movies_plot<- ggplot(movies, aes(x = factor(title_year), y = imdb_score)) +
  geom_boxplot(fill = "#4271AE", colour = "#1F3552") +
  theme(legend.position = "none") +
  theme(axis.text.x = element_text(hjust = 1)) + 
  scale_y_continuous(breaks = seq(0, 10, 1), name = "IMDB Rating") +
  scale_x_discrete(breaks = seq(1980, 2016, 2), name = "Year") + 
  ggtitle("IMDB Ratings Over the Decades") +
  theme_bw() + stat_boxplot(geom = 'errorbar') +
  geom_hline(aes(yintercept = modern_avg), colour = "#D55E00") +
  geom_hline(aes(yintercept = modern_avg, colour = "Average IMDB Rating from 1980-2016"), show.legend = TRUE) +
  theme(legend.position = "bottom")
movies_plot
```

```{r}
stats
```

From the looks of it, the 1980s was overall perhaps the best decade for films, followed by the 1990s. The 80s had the highest quartiles and the highest average score. On the other hand, the 90s had the best film across not only the post 1980-era but of all modern film (*The Shawshank Redemption*), and the "best worst" film with *Police Academy: Mission to Moscow* with a 3.3 rating.

For most of the decade, films on average rated around ~7 or ~8, with few if any lower outliers. Conversely, the 2000s have a number of uncharacteristically bad films which drag down the averages per year quite a bit. One wonders how the 2000s would be seen in terms of film history without these.

Several directors came out strong not only in individual decades, but across them. Christopher Nolan had as many as 4 between 2000 and 2016, with 3 in one decade, and 2 from his Dark Knight Trilogy. Robert Zemeckis and Stephen Spielberg each had one film per decade, and Peter Jackson's Lord of the Rings Trilogy took three of the top spots in the decade it was released.

On the other hand, Jason Friedberg was possibly the most panned director by audiences, given that he released two of the worst films of the 2000s as well as one in the 2010s as voted by IMDB users. Having said that, his "___ Movie" spoof series tend to do well at the box office in spite of being panned by both critics and audiences.

Several of the worst films were sequels to well-known and successful films including horror franchises like *Friday the 13th*, *Halloween*, comedy films like *The Mask*, or the classic that is *Jaws*. Also important to note is that several parody films, including those by Jason Friedberg and the Scary Movie franchise, appeared in the worst film lists. 

## Outstanding Questions

It would be interesting to see how IMDB user rating correlates with a film's gross, or how the number of faces on a poster (one of the metrics in the dataset) correlates with the film's rating. Genre may also be something worth scrutinizing (e.g. what genres were most popular in what decades). 

Some important things to keep in mind when considering the conclusions of this study is that the sample size in the recent decades was much larger than in the earlier ones, so there was a lot more data to substantiate their ratings. i.e. A person is allowed to question whether or not the average ratings for the earlier decades are truly statistically significant, and this is something they might explore in their own analysis.

----
Data found on [Kaggle.com](https://www.kaggle.com/deepmatrix/imdb-5000-movie-dataset) and drawn from the [Internet Movie Database](imdb.com)
